---
title: "Lab 1 Returns Calculator"
author: "T.A. Meraxa"
date: "October 23, 2015"
output: html_document
---
***
### 1 Load the monthly Starbucks return data

In this first lab, you will analyze the monthly stock returns of Starbucks (ticker: SBUX).
Let us get started by downloading the monthly return data from http://assets.datacamp.com/course/compfin/sbuxPrices.csv, and by using the ``read.csv()`` function. Type `?read.table` in the console to consult the help file.
In the ``read.csv()`` function, you should indicate that the data in the CSV file has a ``header`` (header argument) and that strings should not be interpreted as factors (``stringsAsFactors`` argument).

#### Intstructions
* Create a new variable ``data_url`` that contains the URL to the CSV file.
* Create a new variable ``sbux_df`` that contains the data frame with the Starbucks data. ``sbux_df`` should be a data frame object. Data frames are rectangular data objects that contain observations in their rows and variables in their columns.


```{r, eval=FALSE}
# Assign the URL to the CSV file
data_url <- "http://assets.datacamp.com/course/compfin/sbuxPrices.csv"

# Load the data frame using read.csv
sbux_df <- read.csv(data_url, header = T, stringsAsFactors = F)
```

***
### 2 Get a feel for the data

Before you analyze return data, it is a good idea to have (at least) a quick look at the data. R has a number of functions that help you do that:


* The ``str()`` function compactly displays the structure of an R object. It is arguably one of the most useful R functions.
* The ``head()`` and ``tail()`` functions shows you the first and the last part of an R object, respectively.
* The ``class()`` function shows you the class of an R object.

#### Intstructions
* Have a look at the structure of the ``sbux_df`` data.
* Have a look at the first and the last part of the Starbucks prices.
* What is the class of the Date column of the ``sbux_df`` data frame?

```{r, eval=FALSE}
# The sbux_df data frame is already loaded in your work space

# Check the structure of 'sbux_df'
str(sbux_df)

# Check the first and last part of 'sbux_df'
head (sbux_df)
tail (sbux_df)


# Get the class of the Date column of 'sbux_df'
class(sbux_df$Date)
```

***
### 3 Extract the price data

You can use square brackets to extract data from the sbux_df data frame like this ``sbux_df[rows, columns]``. To specify which rows or columns to extract, you have several options:

``sbux_df[1:5, "Adj.Close"]``
``sbux_df[1:5, 2]``
``sbux_df$Adj.Close[1:5]``.
These expressions will all extract the first five closing prices. If you do not provide anything for the rows (or columns), all rows (or columns) will be selected (e.g. ``sbux_df[,"Adj.Close"]``). Check this yourself by typing the different options in the console!
Note that in the above operations, the dimension information was lost. To preserve the dimension information, add the ``drop=FALSE`` argument.

#### Instructions
* Assign to the variable ``closing_prices`` all the adjusted closing prices while preserving the dimension information.

```{r, eval=FALSE}
# The sbux_df data frame is already loaded in your work space
sbux_df[1:5, "Adj.Close"]
sbux_df[1:5, 2]
sbux_df$Adj.Close[1:5]

closing_prices = sbux_df[,2, drop = F]
```

***
### 4 Find indices associated with the dates 3/1/1994 and 3/1/1995

It will often be useful to select stock data between certain dates. Advanced users are advised to look at the xts (http://www.rdocumentation.org/packages/xts) package. However, base R also provides sufficient functionality to do this.
The ``which()`` function returns the indices for which a condition is ``TRUE``. For example: ``which(sbux_df$Date == "3/1/1994")`` returns the position of the date 3/1/1994, which indicates in this case the row number in the ```` data frame.

#### Instructions
* Assign to ``index_1`` the row number of the row that contains the price on 3/1/1994.
* Assign to ``index_2`` the row number of the row that contains the price on 3/1/1995.
* Assign to ``some_prices`` the SBUX closing prices between 3/1/1994 and 3/1/1995.

```{r, eval=FALSE}
# The sbux_df data frame is already loaded in your work space

# Find indices associated with the dates 3/1/1994 and 3/1/1995
index_1 <- which(sbux_df$Date == "3/1/1994")
index_2 <- which(sbux_df$Date == "3/1/1995")

# Extract prices between 3/1/1994 and 3/1/1995
some_prices <- sbux_df[index_1:index_2, "Adj.Close"]

```

***
### 5 Subset directly on dates

The way you selected the data from a specific trading day in the previous exercise was not very convenient, right?
When you create a data frame that has the dates of the stock price as row names, you can select the price on a specific day much more easily. The sample code on the right creates a new data frame ``sbux_prices_df`` that has the trading days as row names. You can select the price on 3/1/1994 now simply with ``sbux_prices_df["3/1/1994", 1]``.

####  Instructions
* Assign to price_1 the Starbucks stock price on 3/1/1994.
* Assign to price_2 the Starbucks stock price on 3/1/1995.

```{r, eval=FALSE}
# The sbux_df data frame is already loaded in your work space

# Create a new data frame that contains the price data with the dates as the row names
sbux_prices_df <- sbux_df[, "Adj.Close", drop=FALSE]
rownames(sbux_prices_df) <- sbux_df$Date
head(sbux_prices_df)
                       
# With Dates as rownames, you can subset directly on the dates.
# Find indices associated with the dates 3/1/1994 and 3/1/1995.
price_1 <- sbux_prices_df["3/1/1994", 1]
price_2 <- sbux_prices_df["3/1/1995", 1]
```

***
### 5 Plot the price data
R has powerful graphical capabilities. On the right, the Starbucks closing prices are plotted as a function of time. This plot was generated with `plot(sbux_df$Adj.Close)`, the basic plotting function (more info in the documentation of plot).
However, we should be able to generate a nicer plot, right? For one thing, a line plot makes much more sense for price time series data.

#### Instructions
* Let us make a better plot by adding the following arguments to the plot function: `type="l"` specifies a line plot, `col="blue"` indicates that the line should be blue, `lwd=2` doubles the line thickness, `ylab="Adjusted close"` adds a y-axis label and `main="Monthly closing price of SBUX"` adds a title.
```{r, eval=FALSE}
# Now add all relevant arguments to the plot function below to get a nicer plot
plot(sbux_df$Adj.Close,
    type = "l", 
    col = "blue",
    lwd = 2,
    ylab = "Adjusted close",
    main = "Monthly closing price of SBUX")
legend(x='topleft',legend='SBUX', lty=1, lwd=2, col='blue')
```

***
### 6 Calculate simple returns
If you denote by $P_{t}$ the stock price at the end of month $t$, the simple return is given by:

$$R_{t} = \frac{P_{t}-P_{t-1}}{P_{t-1}}$$,

the percentage price difference.


Your task in this exercise is to compute the simple returns for every time point n. The fact that R is vectorized makes that relatively easy. In case you would like to calculate the price difference over time, you can use


`sbux_prices_df[2:n,1] - sbux_prices_df[1:(n-1),1]`

Think about why this indeed calculates the price difference for all time periods. The first vector contains all prices, except the price on the first day. The second vector contains all prices except the price on the last day. Given the fact that R takes the element-wise difference of these vectors, you get ${P_{t}-P_{t-1}}$ for every $t$.

#### Instructions
* Assign to sbux_ret the simple returns for the Starbucks stock.
```{r, eval=FALSE}
# The sbux_df data frame is already loaded in your work space
sbux_prices_df <- sbux_df[, "Adj.Close", drop=FALSE]

# Denote n the number of time periods
n <- nrow(sbux_prices_df)
sbux_ret <- (sbux_prices_df[2:n,1] - sbux_prices_df[1:(n-1),1])/sbux_prices_df[1:(n-1),1]

# Notice that sbux_ret is not a data frame object
class(sbux_ret)
sbux_ret
```


***
### 7 Add dates to simple return vector

The vector `sbux_ret` now contains the simple returns of Starbucks (well done!). It would be convenient to have the dates as names for the elements of that vector. Remember that the trading dates were in the first column of the `sbux_df` data frame. To set the names of a vector, you can use `names(vector) <- some_names`.

Remember that we are dealing with closing prices. The first return in sbux_df is thus realized on the second day, or `sbux_prices_df[2,1]`.

#### Instructions
* Assign the correct dates as names to all elements of the return vector `sbux_ret`.
* Use the `head()` function to display the first elements of sbux_ret.

```{r, eval=FALSE}
# The sbux_df data frame is already loaded in your work space
sbux_prices_df <- sbux_df[, "Adj.Close", drop=FALSE]

# Denote n the number of time periods:
n <- nrow(sbux_prices_df)
sbux_ret <- ((sbux_prices_df[2:n, 1] - sbux_prices_df[1:(n-1), 1])/sbux_prices_df[1:(n-1), 1])

# Notice that sbux_ret is not a data frame object
class(sbux_ret)

# Now add dates as names to the vector and print the first elements of sbux_ret to the console to check
names(sbux_ret) <- sbux_df [2:n, 1]
head(sbux_ret)
sbux_ret
```

***
### 8 Compute continuously compounded 1-month returns

As you might remember from class, the relation between single-period and multi-period returns is multiplicative for single returns. That is not very convenient. The yearly return is for example the geometric average of the monthly returns.

Therefore, in practice you will often use continuously compounded returns. These returns have an additive relationship between single and multi-period returns and are defined as

$$Rt = ln(1+R_{t})$$

with $Rt$ the simple return and $rt$ the continuously compounded return at moment $t$.
Continuously compounded returns can be computed easily in R by realizing that

$$Rt = ln(P_{t})-ln(P_{t-1})$$

In R, the log price can be easily computed through `log(price)`. Notice how the `log()` function in R actually computes the natural logarithm.

#### Instructions
* Compute the continuously compounded returns and assign to `sbux_ccret`.
* Set the names of the `sbux_ccret` to the correct dates, as you did in the previous exercise.
* Use the `head()` function to display the first elements of `sbux_ccret`.

```{r, eval=FALSE}
# The sbux_df data frame is already loaded in your work space
sbux_prices_df <- sbux_df[, "Adj.Close", drop=FALSE]

# Denote n the number of time periods:
n <- nrow(sbux_prices_df)
sbux_ret <- ((sbux_prices_df[2:n, 1] - sbux_prices_df[1:(n-1), 1])/sbux_prices_df[1:(n-1), 1])

# Compute continuously compounded 1-month returns
sbux_ccret <- log(sbux_prices_df[2:n, 1]) - log(sbux_prices_df[1:(n-1), 1])

# Assign names to the continuously compounded 1-month returns
names(sbux_ccret) <- sbux_df[2:n,1]

# Show sbux_ccret
head(sbux_ccret)
```


***
### 9 Compare simple and continuously compounded returns

You would like to compare the simple and the continuously compounded returns. In the next exercise, you will do that by generating two graphs. In this exercise, you will just have a quick look at the data. It would be nice to have the simple and continuously compounded return next to each other in a matrix, with n rows and two columns. You can use the `cbind()` function to paste the two vectors that contain both types of returns next to each other in a matrix.

#### Instructions
* Use the `cbind()` and `head()` functions to display the simple and continuously compounded returns next to each other for the first time periods. What do you notice?

```{r, eval=FALSE}
# The sbux_df data frame is already loaded in your work space
sbux_prices_df <- sbux_df[, "Adj.Close", drop=FALSE]

# Denote n the number of time periods:
n <- nrow(sbux_prices_df)
sbux_ret <- ((sbux_prices_df[2:n, 1] - sbux_prices_df[1:(n-1), 1])/sbux_prices_df[1:(n-1), 1])

# Compute continuously compounded 1-month returns
sbux_ccret <- log(sbux_prices_df[2:n,1]) - log(sbux_prices_df[1:(n-1),1])
names(sbux_ccret) <- sbux_df[2:n,1]
head(sbux_ccret)

# Compare the simple and cc returns
head(cbind(sbux_ccret, sbux_ret))
```

***
### 10 Graphically compare the simple and continuously compounded returns

In this exercise, you will create a plot that contains both the simple and continuously compounded returns. This makes it easy to compare both types of returns.

Have a look at the sample code on the right. First of all, we have to plot the simple returns as a function of time. The argument type=l specifies a line plot, col=blue specifies that the simple returns line is blue, lwd=2 specifies the line thickness, ylab="Return" specifies that "Return" is the label of the y-axis and main specifies the plot's main title.

####  Instructions
* Your task is to add the continuously compounded returns to the plot. This can be done with the `lines()` function, having `sbux_ccret` as a first argument. Furthermore, the line should be red and its thickness should be set to 2. You can do that by setting the `col` and `lwd` arguments.

```{r, eval=FALSE}
# The simple returns (sbux_ret) and the continuously compounded returns (sbux_ccret) have been preloaded in your workspace

# Plot the returns on the same graph
plot(sbux_ret, type="l", col="blue", lwd=2, ylab="Return",
               main="Monthly Returns on SBUX")

# Add horizontal line at zero
abline(h=0)

# Add a legend
legend(x="bottomright", legend=c("Simple", "CC"), 
       lty=1, lwd=2, col=c("blue","red"))

# Add the continuously compounded returns
lines(sbux_ccret,
    col = "red",
    lwd = 2)
```
