---
title: "Lab 2 Random Variables and Probabilities"
author: "T.A. Meraxa"
date: "September, 2015"
output: html_document
---

***
### 1. Compute probabilities

In this lab you will become more familiar with random variables and probability distributions.
Suppose $X$ is a normally distributed random variable with a mean $0.05$ and a variance $(0.10)^2$. Start by computing some probabilities.
Probabilities for normally distributed random variables can be computed with the function `pnorm()`. For instance, if $Z$ is a standard normal random variable, $Pr(Z \le 0.5)$ can be computed with `pnorm(0.5)`. You can supply the mean and the standard deviation of the normal distribution via the arguments mean and sd, respectively.

#### Intstructions
* Assign to mu_x the mean of X.
* Assign to sigma_x the standard deviation of X.
* Compute the following probabilities and print the result to the console:
    + Pr(X > 0.10)
    + Pr(X < -0.10)
    + Pr(-0.05 < X < 0.15)

```{r, eval=FALSE}
# X ~ N(0.05, (0.10)^2)
mu_x = 0.05
sigma_x = 0.10 

# Pr(X > 0.10)
1 - pnorm(.10, mean=mu_x, sd=sigma_x)

# Pr(X < -0.10)
pnorm(-.10, mean=mu_x, sd=sigma_x)

# Pr(-0.05 < X < 0.15)
x = pnorm(-0.05, mean=mu_x, sd=sigma_x)
y = pnorm(0.15, mean=mu_x, sd=sigma_x)
y-x
```

***
### 2. Compute quantiles

Continue with computing quantiles of the normally distributed random variable $X$ with a mean $0.05$ and a variance $(0.10)^2$.

Quantiles of normally distributed random variables can be computed with the function `qnorm()`. For instance, if `Z` is a standard normal random variable, the 10% quantile can be computed with `qnorm(0.1)`. You can supply the mean and the standard deviation of the normal distribution via the arguments `mean` and `sd`, respectively.

#### Instructions
* Compute the 1%, 5%, 95% and 99% quantile with a single command. Do not assign it to a variable, just print it to the console.


```{r, eval=FALSE}
# 1%, 5%, 95% and 99% quantile
qnorm(c(0.01, 0.05, 0.95, 0.99), mean = mu_x, sd = sigma_x)
```

***
### 3. Compute densities

Now consider two stocks: Microsoft (ticker: MSFT) and Starbucks (ticker: SBUX). Let $X$ denote the monthly return on the Microsoft stock and let $Y$ denote the monthly return on the Starbucks stock. Assume that $X$ is normally distributed with a mean $0.05$ and a variance $(0.10)^2$, and that $Y$ is normally distributed with a mean $0.025$ and a variance $(0.05)^2$.

Values of a normal density can be computed with function `dnorm()`. For instance, the value of the standard normal density at 0 can be computed with  `dnorm(0)`. You can supply the mean and the standard deviation of the normal distribution via the arguments `mean` and `sd`, respectively.

#### Instructions
* For the grid of values in `x_vals`, compute the values of the assumed normal density for the return on Microsoft stock and assign it to `MSFT`.
* For the grid of values in `x_vals`, compute the values of the assumed normal density for the return on Starbucks stock and assign it to `SBUX`.

```{r, eval=FALSE}
# Normally distributed monthly returns
x_vals <- seq(-0.25, 0.35, length.out = 100)
MSFT <- dnorm(x_vals, mean = 0.05, sd = 0.10)
SBUX <- dnorm(x_vals, mean = 0.025, sd = 0.05)
```

***
### 4. Plot normal curve

Continue with the two stocks. The aim is to produce a single plot containing the two assumed normal curves. Start with plotting the normal curve for the return on Microsoft stock.

With `plot(x, y)`, you can plot the values of `y` on the y-axis against the values of `x` on the x-axis. You can customize the plot in many ways through additional arguments (http://www.rdocumentation.org/packages/graphics/functions/plot.default): set `type = "l"` to plot the data as a connected line, use `col` to specify the line color, set axis labels with `xlab` and `ylab`, or set axis limits with  `xlim` and `ylim`.

#### Instructions
* Plot a line for the values of the assumed normal density for the return on Microsoft stock (MSFT) against `x_vals`.
* Customize the plot in the following way: use a blue line, set the y-axis label to `"Normal curves"`, and set the y-axis to be in the interval $[0,8]$.


```{r, eval=FALSE}
# MSFT and x_vals are still in your workspace

# Normal curve for MSFT
plot (x_vals, MSFT,
    type = "l",
    lwd = 2,
    col  = "blue",
    ylab = "Normal curves",
    xlim=c(-1, 1), 
    ylim=c(0, 8)
    )
```

***
### 5. Add second normal curve
Now add the normal curve for the return on Starbucks stock to the plot.

You can add lines to a plot with `lines(x, y)`. Furthermore, you can specify the line color with the `col` argument.

Note that a plot legend is added for you in the code on the right.

#### Instructions
* Add a red line for the values of the assumed normal density for the return on Starbucks stock (`SBUX`) against `x_vals`.
```{r, eval=FALSE}
# MSFT, SBUX and x_vals are still in your workspace

# Normal curve for MSFT
plot(x_vals, MSFT, type = "l", col = "blue", ylab = "Normal curves", 
     ylim = c(0, 8))

# Add a normal curve for SBUX
lines(x_vals, SBUX, tyle = "l", col = "red")

# Add a plot legend
legend("topleft", legend = c("Microsoft", "Starbucks"), 
       col = c("blue", "red"), lty = 1)
```

***
### 6. Determine the value-at-risk of simple monthly returns

Consider again the Microsoft stock. Assume that the simple monthly return $R$ is normally distributed with a mean $0.04$ and a variance $(0.09)^2$. The initial wealth to be invested over the month is $100,000.

Determine the 1% and the 5% value-at-risk (VaR) over the month on the investment. That is, determine the loss in investment value that may occur over the next month with a 1% probability and with a 5% probability.

#### Instructions
* Assign to `mu_R` the mean of R.
* Assign to `sigma_R` the standard deviation of R.
* Assign to `W0` the initial wealth.
* Compute the 1% value-at-risk and print the result to the console.
* Compute the 5% value-at-risk and print the result to the console.

```{r, eval=FALSE}
# R ~ N(0.04, (0.09)^2) 
mu_R <- 0.04
sigma_R <- 0.09

# Initial wealth W0 equals $100,000
W0 <- 100000

# The 1% value-at-risk
W0 * qnorm(0.01, mean = mu_R, sd = sigma_R)

# The 5% value-at-risk
W0 * qnorm(0.05, mean = mu_R, sd = sigma_R)
```

***
### 7. Determine the value-at-risk of continuously compounded monthly returns

Instead of the simple monthly return, now look at the continuously compounded monthly return r of the Microsoft stock. Assume that $r$ is normally distributed with a mean $0.04$ and a variance $(0.09)^2$. The initial wealth to be invested over the month is $100,000.

Determine the 1% and the 5% value-at-risk (VaR) over the month on the investment. That is, determine the loss in investment value that may occur over the next month with a 1% probability and with a 5% probability.

Use the fact that the continuously compounded return quantile can be transformed to a simple return quantile with the transformation $R=e^r−1$. The exponential $e^r$ can easily be computed with `exp(r)`.

#### Instructions
* Assign to `mu_r` the mean of r.
* Assign to `sigma_r` the standard deviation of r.
* Assign to `W0` the initial wealth.
* Compute the 1% value-at-risk and print the result to the console.
* Compute the 5% value-at-risk and print the result to the console.
```{r, eval=FALSE}
# r ~ N(0.04, (0.09)^2) 
mu_r <- 0.04
sigma_r <- 0.09

# Initial wealth W0 equals $100,000
W0 <- 100000

# The 1% value-at-risk
W0 * (exp (qnorm(0.01, mean = mu_r, sd = sigma_r)) -1)

# The 5% value-at-risk
W0 * (exp (qnorm(0.05, mean = mu_r, sd = sigma_r)) -1)
```

***
### 8. Compute simple monthly returns

For the remaining exercises, consider a one-month investment in two Northwest stocks: Amazon and Costco.
Suppose you buy Amazon and Costco at the end of September at $P_{A,t−1}=38.23$,$P{C,t−1}=41.11$ and then sell at the end of October for $P_{A,t}=41.29$,$P_{C,t}=41.74$. Compute the simple monthly returns for the two stocks.

Note that the above prices were the actual closing prices for 2004 taken from Yahoo!.


#### Instructions
Assign to `PA` a vector that contains the prices of the Amazon stock.
Assign to `PC` a vector that contains the prices of the Costco stock.
Compute the simple monthly return on the Amazon stock and assign it to `RA`.
Compute the simple monthly return on the Costco stock and assign it to `RC`.

```{r, eval=FALSE}
# Vectors of prices
PA <- c(38.23,41.29)
PC <- c(41.11,41.74)

PA[2]

# Simple monthly returns
RA <- (PA[2] - PA[1])/PA [1]
RC <- (PC[2] - PC[1])/PC [1]

RA
RC
```

***
### 9. Compute continuously compounded monthly returns

Now compute the continuously compounded monthly returns on the Amazon and Costco stocks.

#### Instructions
Compute the continuously compounded monthly return for Amazon and assign it to `rA`.
Compute the continuously compounded monthly return for Costco and assign it to `rC`.

```{r, eval=FALSE}
# Continuously compounded returns
rA <- log(RA +1)
rC <- log(RC +1)

rA
rC
```

***
### 10. Compute simple total returns and dividend yields

Suppose Amazon paid a $0.10 per share cash dividend at the end of October. Compute the monthly simple total return on Amazon and the monthly dividend yield.

#### Instructions
* Assign to `DA` the cash dividend per share.
* Compute the monthly simple total return and assign it to `RA_total`.
* Compute the monthly dividend yield and assign it to `DY`.

```{r, eval=FALSE}
# The prices for Amazon (PA) are still in your workspace

# Cash dividend per share
DA <- 0.10

# Simple total return
RA_total <- (PA[2]-PA[1]+DA)/PA[1]

# Dividend yield
DY <- DA/PA[1]

RA_total
DY
```

***
### 11. Compute annual returns

Suppose the monthly returns on Amazon are equal to `RA` every month for one year. Compute the simple annual returns, as well as the continuously compounded annual returns.

#### Instructions
* Compute the simple annual return and assign it to `RA_annual`.
* Compute the continuously compounded annual return and assign it to `rA_annual`.

```{r, eval=FALSE}
# The simple monthly return on Amazon (RA) is still in your workspace

# Simple annual return
RA_annual <- ((1 + RA))^12-1

# Continuously compounded annual return
rA_annual <- log(1 + RA_annual)

RA_annual
rA_annual
```

***
### 12. Compute portfolio shares and portfolio returns

At the end of September 2004, suppose you have $10,000 to invest in Amazon and Costco over the next month.

Compute your portfolio shares if you invest $8,000 in Amazon and $2,000 in Costco. Moreover, compute the simple monthly return on the portfolio and assume that no dividends were paid.

#### Instructions
* Compute the portfolio share of your Amazon stocks and assign it to xA.
* Compute the portfolio share of your Costco stocks and assign it to xC.
* Compute the simple monthly return on the portfolio and print the result to the console.

```{r, eval=FALSE}
# The simple returns on Amazon (RA) and Costco (RC) are still in your workspace

# Portfolio shares
xA <- 8000/10000
xC <- 2000/10000

# Simple monthly return
xA*RA + xC*RC
```
