---
title: "Lab 3 Bivariate Distributions"
author: "T.A. Meraxa"
date: "September, 2015"
output: html_document
---

***
### 1. Covariance matrix

In this lab you will learn to draw observations from bivariate normal distributions, and explore the resulting data to familiarize yourself with relationships between two random variables.

Let us assume that $X$ and $Y$ are bivariate normally distributed. Get started by defining a covariance matrix. Suppose that the standard deviations are $σ_X=0.10$ and $σ_Y=0.05$, and that the correlation is $ρ_XY=0.9$.

Matrices can be created with the `matrix()` function. You should supply the values as the first argument and specify the dimensions with the `nrow` and  `ncol` arguments. For example, `matrix(c(-0.5, 0.5), nrow = 1, ncol = 2)` creates a matrix with one row and two columns. By default, the matrix is filled column by column, but you can fill the matrix row by row by setting  `byrow = TRUE`.

#### Instructions
* Compute the covariance between X and Y and assign it to sig_xy.
* Construct the covariance matrix and assign it to Sigma_xy.
```{r, eval=FALSE}
# Standard deviations and correlation
sig_x <- 0.10
sig_y <- 0.05
rho_xy <- 0.9

# Covariance between X and Y
sig_xy <- rho_xy*sig_x*sig_y 

# Covariance matrix
Sigma_xy <-  matrix(c(sig_x^2, sig_xy, sig_xy, sig_y^2), 2, 2, byrow=TRUE)
```

***
### 2. Simulate data

Let us continue by simulating observations from the bivariate normally distributed random variables $X$ and $Y$.

Notice that the second line in the R script on the right loads the mvtnorm R package. `mvtnorm` is an R library, containing extra functionality (see mvtnorm documentation), which makes it easy to simulate data from a bivariate (or more generally, a multivariate) normal distribution. We will be using the `rmvnorm()` function from the package. If you would like to use the `mvtnorm` package at home, use the `install.packages` function to first install the package on your own computer.

You can supply the mean vector and the covariance matrix via the arguments  `mean` and `sigma`, of the `rmvnorm()` function respectively. For instance, you can draw 10 observations from a bivariate standard normal distribution with `rmvnorm(10, mean = c(0, 0), sigma = diag(2))`.

When you simulate data, it is often useful to have *reproducible results*. This can be achieved by setting the seed of the random number generator beforehand with the `set.seed()` function. For more information, see the documentation on random number generation.

####  Instructions
* Simulate 100 observations from the bivariate normal distributed random variables $X$ and $Y$. Use the covariance matrix `Sigma_xy` from the previous exercise and suppose that the means are $μ_X=0.05$ and $μ_Y=0.025$. Assign the result to `xy_vals`.
* Have a look at the first few observations with the `head()` function.

```{r, eval=FALSE}
# Load the mvtnorm package
library("mvtnorm")

# The covariance matrix (Sigma_xy) is still in your workspace

# Means
mu_x <- 0.05
mu_y <- 0.025

# Simulate 100 observations
set.seed(123)  # for reproducibility
xy_vals <- rmvnorm(100, mean = c(mu_x, mu_y), sigma = Sigma_xy)
  
# Have a look at the first observations
head(xy_vals)
```


